An Analysis of Performance Bottlenecks in MRI Pre-Processing

### 2024-02-13 paper/paper.tex
[WIP]

\usepackage{cite}
\usepackage{xcolor}
\usepackage{hyperref}
\hypersetup{colorlinks=true,linkcolor=black,citecolor=blue,filecolor=black,urlcolor=blue}
\usepackage{graphicx}
\graphicspath{{figures}}
\usepackage{caption}
\usepackage{subcaption}
\captionsetup[subfigure]{singlelinecheck=false, skip=0pt}
\usepackage{listings}
\renewcommand{\lstlistingname}{Algorithm}
\usepackage[binary-units=true]{siunitx}
\usepackage{csvsimple}
\usepackage{booktabs}

@@

\title{A Detailed Analysis of Performance Bottlenecks in MRI Pre-Processing}

\author{Mathieu Dugr\'e, Tristan Glatard \TG{I think Yohan should be co-author given his technical input}}

\begin{document}

\maketitle

@@

\section{Introduction}
% Intro: Large dataset, long processing time, I/O intensive (derivatives), focus on MRI, no current study, pipelines?
% **Main question:** What are the main bottleneck in MRI pre-processing?
% **Why:** Performance of neuroimaging pipeline is critical (larger study, clinical use). However, limited study on performance profile.

\TG{Good intro but I think it would deserve a better logic. You could start by explaining that computational bottlenecks 
have been one of the main bottlenecks to the processing of large cohorts in neuroimaging, that the community is constantly searching for ways to process data faster (HPC, cloud, GPUs, DL models), and that your paper takes a step back by analyzing the computational bottlenecks in order to subsequently focus on their optimization. 
}

Magnetic Resonance Imaging (MRI) is a standard tool used by neuroscientists to perform clinical diagnosis and for researchers to develop a better understanding of the brain. Three main MRI techniques exist: structural MRI (sMRI), functional MRI (fMRI), and diffusion MRI (dMRI). While other modalities such as EEG, CT, and PET exist \TG{define acronyms}, we focus on MRIs for their broad adoption and non-invasive property. However, MRI data analysis is challenging due to computationally expensive pre-processing and large output and intermediate data size.

Neuroscientists developed various toolboxes to tackle the challenging task of pre-processing MRI data. Among those, some openly available and commonly accepted emerged: AFNI~\cite{Cox1996-sl}, ANTs~\cite{Avants_undated-fu}, FreeSurfer~\cite{Fischl2012-bp}, FSL~\cite{Jenkinson2012-cq}, and SPM~\cite{Friston2007-ag}. While the previous pipelines focus on specific steps of MRI pre-processing, tools like fMRIPrep~\cite{Esteban2019-og} and dMRIPrep~\cite{Michael2021-4515513} combine various pipelines to produce a complete pre-processing workflow.

The MRI pre-processing of a subject can take several hours, preventing the clinical use where data analysis might be required in timely manner. Moreover, the lengthy pre-processing hinders researchers with limited computational resources to perform research on large cohorts.

The first step to improve performance of an application is to understand its performance bottleneck. Following, we can focus our efforts on the performance critical components. For example, better algorithm designs or usage of hardware accelerator can reduce computation time. Compression techniques can lower data transfer time for output and intermediate data. At last, reduced precision arithmetic can lower computation time, memory footprint, and storage size. While these techniques can provide significant performance improvements, their implementation can require substantial effort. Therefore, it is critical to understand the performance bottleneck of MRI pre-processing pipelines to improve their performance.

The neuroimaging community is well aware that MRI pre-precessing pipelines require long computation time. However, to the best of our knowledge, there is no comprehensive study of the performance bottlenecks in MRI pre-processing. In this paper, we characterize computational cost of several commonly adopted MRI pre-processing pipelines. The results of our analysis serve as a reference for future efforts to optimize MRI pre-processing workflows.

\TG{you could summarize the method (vtune profiler) in the intro.}

\section{Methods}
\subsection{Infrastructure}
For our experiments, we used the \textit{Slashbin cluster} at Concordia University. The cluster has eight compute nodes, each configured with two 16-core Intel(R) Xeon(R) Gold 6130 CPU @ \SI{2.10}{\giga\hertz}, \SI{250}{\gibi\byte} of RAM, \SI{126}{\gibi\byte} of tmpfs, six \SI{447}{\gibi\byte} SSDs with the XFS file system (no RAID enabled), CentOS~8.1 and Linux kernel \textit{4.18.0-477.10.1.el8\_lustre.x86\_64}. \TG{I would defer this subsection to later in the methods as it's not the most central one}

@@

We used the the OpenNeuro ds004513 v.1.0.2 dataset~\cite{ds004513:1.0.2}. The anatomical, functional, and diffusion data from 20 health individuals was acquired from two different cohorts. The within-subject cohort has nine participants (mean age=43~yrs, std=7~yrs; 4 females) with two session: eyes open and eyes closed. The replication cohort has eleven participants (mean age=27~yrs, std=5~yrs; 6 females) with only the eyes open session. We profiled the application using both cohort on the eyes open session, unless explicitly specified.

\subsection{Pipelines}
% TODO rework
In this analysis, we aim to cover the pre-processing performance for different MRI modalities. We study the performance bottleneck of large toolboxes and individual pipelines. Moreover, we use several pipelines from the LivingPark project~\cite{livingpark} as a real use case of MRI pre-processing.

First, we profile these commonly accepted workflows: fMRIPrep anatomical only without and without FreeSurfer, fMRIPrep complete workflow, and dMRIPrep. This lets us study the coarse grain performance of workflows for anatomical, functional, and diffusion MRI pre-processing.

Second, we perform a finer grain analysis of the aforementioned workflows' sub-components. We list the different sub-workflow and describe the key steps for each modality in the section below.

Last, the LivingPark project replicates several papers on Parkinson's Disease with a broad range of analysis. The study uses the following pipelines: FSL SIENA/VIENA, SPM VBM, SPM DARTEL, and SPM pairwise registration.
% TODO Decide which pipelines to use from the project.

% ############################################################
% Pre-Processing
% ############################################################
% Look at the benchmarked tool and the information needed to abstract those tools.
% Discussion of the pre-processing steps.
% Explain their computation and I/O requirements.
\subsection{sMRI Pre-Processing}
\subsubsection{Intensity correction}
% N4 bias correction (Check ANTS brainExtraction)
% https://ieeexplore.ieee.org/document/5445030

\subsubsection{Reference Image}
% Longitudinal study 
% FreeSurfer mri_robust_template

\subsubsection{Spatial Normalization}
% ANTS registrationSyN
% SyN Registration: https://pubmed.ncbi.nlm.nih.gov/17659998/

\subsubsection{Skull-Stripping}
% ANTS brainExtraction
% 1. Truncate image intensity
% 2. N4 bias correction
% 3. SyN Registration
% 4. Warp prior to image using affine
% 5. Atropos: https://pubmed.ncbi.nlm.nih.gov/21373993/

\subsubsection{Segmentation}
% FSL FAST

\subsubsection{Surface Reconstruction}
% FreeSurfer recon-all

\subsection{fMRI Pre-Processing}
\subsubsection{Slice-Timing Correction}
% AFNI 3dTShift

\subsubsection{Head-Motion Correction}
% FSL MCFLIRT

\subsubsection{Susceptibility Distortion Correction}
% SDC workflow

\subsubsection{Registration}
% FSL FLIRT

\subsection{dMRI Pre-Processing}
% TODO: Decide which pipelines to use
% \subsubsection{Eddy Currents Distortion Estimation}
% % FSL eddy

\subsection{Profiling}
Profiling MRI pre-processing pipelines present a set of challenges. The neuroimaging community develops pipelines using several programming languages including \textit{C}, \textit{C++}, \textit{Fortran}, \textit{Python}, and \textit{Matlab}, with several pipelines utilizing a combination of multiple languages. Pipelines are generally complex and computationally expensive.

The Intel VTune profiler~\cite{vtune_profiler} satisfies these challenges by offering multi-language support, low performance overhead, and information of functions at runtime level. However, ome general profiling challenges remain. Profilers require source code to be compiled with Debug Symbols to report useful information about the different functions and modules. Moreover, different input data can substantially vary the analysis due to conditional branching in the application and convergence thresholds. We discuss our approach to these challenges in the next section.
% TODO add citation on effect of different input data on numerical stability and reduced precision.

First, to obtain human readable information on function and module names in VTune, we re-compiled each pipeline with debug information in a Docker image. For use on HPC systems, we created Singularity images using Docker2Singularity \lstlistingname~\ref{vtune_example} illustrate the profiling of a pipeline by mounting the VTune binary to the Singularity image at execution.
\lstinputlisting[
	language={sh},
	caption={Profiling an application within a Singularity container, using VTune profiler.},
	captionpos=b,
	label=vtune_example,
	numbers=left,
	frame=single,
	basicstyle=\footnotesize,
	numbersep=5pt,
	numberstyle=\tiny\color{gray},
	rulecolor=\color{black},
	tabsize=2,
]{vtune-example.sh}

Second, to estimate an average performance profile, we aggregate the profiling results of each pipeline across all subjects. Combining the Pareto's law and VTune hotspots report \TG{It would be useful to define these terms}, we present the functions accounting for most of the CPU time up to 80\% of the total CPU time of the application \TG{check wording}. This heuristic allows us to focus our efforts on the performance critical sections of a pipeline. We report the mean and standard deviation of the execution time for those functions.

Last, we profile the application two threading configuration: single-threaded and 32 threads. On the one hand, the single-threaded approach limits the potential overhead from I/O on the profiling. On the other hand, we evaluate the multi-threading performance of the applications using maximum number of CPU core available on our system (i.e. 32).

@@

\section{Results}
\TG{Interesting results but presentation needs to be worked out. This section should be structured with sub-headings summarizing the main findings, for instance: "Interpolation is the main bottleneck", etc}

% TODO
% 2-step approach:
% 1. Which pipeline constitute the long execution time in preprocessing?
% 2. For those, which functions take the most time to execute? (hotspots)
% Any key finding: common point between functions, mem bound, fops, etc.

% In this section, analyze the aggregated profiling data obtained with VTune. First, a coarse summary of the workflows and pipelines is presented using their makespan. Second, we depict detailed analysis for the key functions contributing to the CPU time of each application (hotspots).

\subsection{Makespan}
% Identify which tools take the longest to execute.
% This will let us decided which pipeline to investigate more.
Figure~\ref{fig:makespan-1thread} shows the mean and standard deviation of the makespan \TG{define makespan}, in seconds, while using 1 thread for one execution of all subjects.
Unexpectedly, we observe that both ANTS~BrainExtration and ANTS~RegistrationSyN are slower when using single precision compared to double precision.
% TODO Stat test between mean for single and double precision.
\begin{figure}[h]
	\centering
	\includegraphics[width=\linewidth]{figures/makespan-1thread.png}
	\caption{Profiling of the pipelines using 1 thread. \TG{Font size in the figure is too small, it should be similar to the caption font size}}
	\label{fig:makespan-1thread}
\end{figure}

\subsection{Hotspots}

Figure~\ref{fig:hotspots-fmriprep-anat-subworkflow} depicts the hotspots observed from profiling the fMRIPrep anatomical sub-workflows. In each sub-figure, the x-axis shows the function ID grouped by parent module, which are ordered by CPU time of (1) the modules, then (2) the functions within a module. The function ID are found in the supplementary material Section~\ref{sec:supplementary}. The bars show the average CPU time in seconds of a function and the error bars indicate the standard deviation, with the left y-axis showing the scale. The light-green dots represent the functions cumulative percentage makespan of the application, with the right y-axis showing the scale. The color of the bar indicate the percentage of calls, for the function, that were ``stalled due to demand memory loads and stores''\footnote{\href{https://www.intel.com/content/www/us/en/docs/vtune-profiler/user-guide/2023-0/cpu-metrics-reference.html\#MEMORY-BOUND}{Vtune Memory Bound metric}}, with the left color bar showing the scale.

\TG{I'd move the graphical explanations to the figure caption so that the text can focus on observations}

In figure~\ref{subfig:hotspots-ants-brainExtraction}, we observe that the function 1 and 2 account for a significant portion of the application execution time. Moreover, calls from function~1 are memory bound $51.84\pm{4.55}$ percent of the time. Table~\ref{extab:hotspots-1thread-ants-brainExtraction} shows that both functions are part of the ITK linear interpolation module.

Similarly, figure~\ref{subfig:hotspots-ants-brainExtraction-fp} shows the same two functions accounting for the most time, although in flipped order. Calls from function~2 are memory bound $36.30\pm{4.98}$ percent of the time. Compared to the double precision version of ANTS brainExtraction, the third function runtime is significantly larger. It is an ITK function transforming points while calculating the displacement field of the image.

Figure~\ref{subfig:hotspots-ants-registrationSyN} and figure~\ref{subfig:hotspots-ants-registrationSyN-fp} show that the execution time for ANTS registrationSyN is dominated by a single function. In both case, the ITK function evaluating point during linear interpolation. The next two function are also shared between the application, although in flipped order. These functions show some memory bound in the range of 16--27.5. We also note the three most expensive functions are the same as in both versions of ANTS brainExtraction.

% TODO FreeSurfer 1-thread sub-figure

In figure~\ref{subfig:hotspots-32threads-freesurfer-reconall}, we observe that OMP is a major bottleneck for the application, accounting for at least 76\% of the application runtime. The only function shown from FreeSurfer recon-all is the same as the first function from figure~\ref{subfig:hotspots-1thread-freesurfer-reconall}. However, it only accounts for 2\% of the total makespan compared to 3.8\% while running sequentially.

\begin{figure*}[ht!]
	\centering
	% ANTS brainExtraction
	\begin{subfigure}[t]{0.49\textwidth}
		\caption{}
		\label{subfig:hotspots-ants-brainExtraction}
		\includegraphics[width=\textwidth]{figures/hotspots-1thread-ants-brainExtraction.png}
	\end{subfigure}
	\hfill
	\begin{subfigure}[t]{0.49\textwidth}
		\caption{}
		\label{subfig:hotspots-ants-brainExtraction-fp}
		\includegraphics[width=\textwidth]{figures/hotspots-1thread-ants-brainExtraction-fp.png}
	\end{subfigure}
	% ANTS registrationSyN
	\begin{subfigure}[t]{0.49\textwidth}
		\caption{}
		\label{subfig:hotspots-ants-registrationSyN}
		\includegraphics[width=\textwidth]{figures/hotspots-1thread-ants-registrationSyN.png}
	\end{subfigure}
	\hfill
	\begin{subfigure}[t]{0.49\textwidth}
		\caption{}
		\label{subfig:hotspots-ants-registrationSyN-fp}
		\includegraphics[width=\textwidth]{figures/hotspots-1thread-ants-registrationSyN-fp.png}
	\end{subfigure}
	\caption{ANTS: Single vs. Double precision}
	\label{fig:ants-fp-comparison}
\end{figure*}

\section{old Results}
% TODO
% 2-step approach:
% 1. Which pipeline constitute the long execution time in preprocessing?
% 2. For those, which functions take the most time to execute? (hotspots)
% Any key finding: common point between functions, mem bound, fops, etc.

% In this section, analyze the aggregated profiling data obtained with VTune. First, a coarse summary of the workflows and pipelines is presented using their makespan. Second, we depict detailed analysis for the key functions contributing to the CPU time of each application (hotspots).

\subsection{Hotspots}
Figure~\ref{fig:hotspots-fmriprep-anat-subworkflow} depicts the hotspot observed from profiling the fMRIPrep anatomical sub-workflows. In each sub-figure, the x-axis shows the function ID grouped by parent module, which are ordered by CPU time of (1) the modules, then (2) the functions within a module. The function ID are found in the supplementary material Section~\ref{sec:supplementary}. The bars show the average CPU time in seconds of a function and the error bars indicate the standard deviation, with the left y-axis showing the scale. The light-green dots represent the functions cumulative percentage makespan of the application, with the right y-axis showing the scale. The color of the bar indicate the percentage of calls, for the function, that were ``stalled due to demand memory loads and stores''\footnote{\href{https://www.intel.com/content/www/us/en/docs/vtune-profiler/user-guide/2023-0/cpu-metrics-reference.html\#MEMORY-BOUND}{Vtune Memory Bound metric}}, with the left color bar showing the scale.

In figure~\ref{subfig:hotspots-ants-brainExtraction}, we observe that the function 1 and 2 account for a significant portion of the application execution time. Moreover, calls from function~1 are memory bound $51.84\pm{4.55}$ percent of the time. Table~\ref{extab:hotspots-1thread-ants-brainExtraction} shows that both functions are part of the ITK linear interpolation module.

Similarly, figure~\ref{subfig:hotspots-ants-brainExtraction-fp} shows the same two function accounting for the most time, although in flipped order. Calls from function~2 are memory bound $36.30\pm{4.98}$ percent of the time. Compared to the double precision version of ANTS brainExtraction, the third function runtime is significantly larger. It is an ITK function transforming points while calculating the displacement field of the image.

Figure~\ref{subfig:hotspots-ants-registrationSyN} and figure~\ref{subfig:hotspots-ants-registrationSyN-fp} show that the execution time for ANTS registrationSyN is dominated by a single function. In both case, the ITK function evaluating point during linear interpolation. The next two function are also shared between the application, although in flipped order. These functions show some memory bound in the range of 16--27.5. We also note the three most expensive functions are the same as in both versions of ANTS brainExtraction.

% TODO FreeSurfer 1-thread sub-figure
% Note:
% - Way more functions
% - Top 5 accounts for only 15% of total time. Much less than other applications.
% - Way more memory bound that previous application.
% - Focusing on Top 5 and memory bound could give good performance boost.

>>>>

\usepackage{cite}
\usepackage{xcolor}
\usepackage{hyperref}
\hypersetup{colorlinks=true,linkcolor=black,citecolor=blue,filecolor=black,urlcolor=blue}
\usepackage{graphicx}
\graphicspath{{figures}}
\usepackage{caption}
\usepackage{subcaption}
\captionsetup[subfigure]{singlelinecheck=false, skip=0pt}
\usepackage{listings}
\renewcommand{\lstlistingname}{Algorithm}
\usepackage[binary-units=true]{siunitx}
\usepackage{csvsimple}
\usepackage{booktabs}
\usepackage{amsmath}
\usepackage{import}

@@

\title{An Analysis of Performance Bottlenecks in MRI Pre-Processing}
\author{Mathieu Dugr\'e, Yohan Chatelain, Tristan Glatard}

\begin{document}
\maketitle

@@

\section{Introduction}
% Computational bottleneck
Pre-processing of Magnetic Resonance Imaging (MRI) data is a critical step in neuroimaging. Although established pre-processing pipelines exist, they commonly require expensive computation and produce large output and intermediate data. This has been a major bottleneck to study large subject cohorts in neuroimaging. Moreover, the execution time of pipelines can hinder clinical applications where timely data analysis is needed. For those reasons, the neuroimaging community is constantly seeking ways to improve the performance of MRI pre-processing pipelines.

% HPCs
Researcher broadly adopted High Performance Computing (HPC) to accelerate the MRI processing of large cohorts. HPC systems provide a large number of CPU cores and high-bandwidth memory to process subjects in parallel. However, HPCs might not be available to researchers in developing country. Similarly, cloud computing allows scaling the number of core and memory to process large cohort. However, cloud computing requires expertise to setup and can be costly for large cohort. Moreover, data transfer can be slow and might be impossible due to ethic concerns related to human data. In both cases, the benefits from parallelization are limited by the parallel efficiency of the pipeline and the number of subjects to process. Therefore, it can lead to under-utilization of the available resources.

%  GPUs
\MD{TODO GPUs / Accelerators}

% DL
In recent years, Deep Learning (DL) has shown promising results in MRI pre-preprocessing. For example, tools such as SynthMorph~\cite{Hoffmann2022-hu}, SynthSeg~\cite{Billot2023-vp}, SynthSR~\cite{Iglesias2023-bp}, and SynthStrip~\cite{Hoopes2022-ms} were integrated into FreeSurfer~\cite{Fischl2012-cx} with the effort to provide Deep Learning alternative to traditional algorithms. The main benefit of those tools is the faster runtime they provide. Moreover, A study~\cite{Pepe2023-dm} showed that the results generated by SynthMorph are numerical more stable than those from FreeSurfer. However, training a DL model requires large amount of data and specialized hardware, and is computationally expensive. Additionally, the results from DL models can be challenging to interpret, explaining the slow adoption of DL in MRI pre-processing.

% Alternative solutions
While those techniques can provide significant performance improvements, their additional hardware requirement can be costly and unaccessible to some researchers. Alternatively, improved algorithm designs, compression techniques, or reduced- and mixed-precision techniques could provide substantial performance improvements. However, their implementation can require substantial effort. Therefore, it is critical to understand the performance bottleneck of MRI pre-processing pipelines to improve their performance.

% - No comprehensive study of the performance bottleneck of MRI pre-processing.
The neuroimaging community is well aware that MRI pre-precessing pipelines require long computation time. However, to the best of our knowledge, there is no comprehensive study of the performance bottlenecks in MRI pre-processing. In this paper, we characterize computational cost of several commonly adopted MRI pre-processing pipelines. The results of our analysis serve as a reference for future efforts to optimize MRI pre-processing workflows.

% - Focus on MRI pre-processing.
Magnetic Resonance Imaging (MRI) is a standard tool used by neuroscientists to perform clinical diagnosis and for researchers to develop a better understanding of the brain. Three main MRI techniques exist: structural MRI (sMRI), functional MRI (fMRI), and diffusion MRI (dMRI). While other modalities such as electroencephalogram (EEG), computed tomography (CT), and positron emission tomography (PET) exist, we focus on MRIs for their broad adoption and non-invasive property. Moreover, MRI data analysis is challenging due to computationally expensive pre-processing and large output and intermediate data size.

% - Pipelines
Neuroscientists developed various toolboxes to tackle the challenging task of pre-processing MRI data. We focus on the commonly accepted fMRIPrep~\cite{Esteban2019-bl} workflow as it a comprehensive pre-processing pipeline for both sMRI and fMRI. fMRIPrep combines several pipelines to produce a complete pre-processing workflow. We also study its sub-workflow separately to provide a finer grain analysis of the performance bottleneck.

% - data: OpenNeuro ds004513
We use the OpenNeuro ds004513 v1.0.2 dataset~\cite{ds004513:1.0.2}. We chose this dataset for its wide range of age and equal distribution of biological sex. Using a diverse dataset should help to prevent findings of potential performance bottleneck associated to a specific population.

% - VTune profiler
To profile the pipelines, we use the Intel VTune profiler~\cite{vtune-profiler}. The VTune profiler is a multi-language profiler with low performance overhead and provides information of functions at runtime level. We profile the application using two threading configuration: single-threaded and 32 threads. We aggregate subjects' results for each pipeline to obtain an average performance profile. We compiled each application with debug information to obtain human readable information on function and module names in VTune.

\section{Materials \& Methods}
\subsection{Pipelines}
In this paper, we study the performance bottleneck of fMRIPrep; a commonly used fMRI and sMRI pre-processing workflow. First, we profile the entire workflow to identify the coarse grain performance bottleneck to obtain a coarse view of the workflow's performance. Following, to characterize the performance at a finer grain, we profile the sub-workflows of fMRIPrep: ANTs brainExtraction, ANTs registrationSyN~\cite{Avants2008-ea}, FreeSurfer recon-all~\cite{Dale1999-wu}, FSL FAST~\cite{Zhang2001-hx}, FSL FLIRT~\cite{Jenkinson2002-od,Jenkinson2001-eu,Greve2009-dw}, and FSL MCFLIRT~\cite{Jenkinson2002-od}. We omitted some sub-workflows of fMRIPrep as they were not compatible with our dataset. Moreover, the statistical analysis of the fMRIPrep workflow are not computationally expensive and are not the focus of this paper.

\subsubsection{ANTS registrationSyN}
% Spatial Normalization
% SyN Registration: https://doi.org/10.1016/j.media.2007.06.004
% TODO Mention the option for single and double precision.

\subsubsection{ANTS brainExtraction}
% ANTS brainExtraction
% 1. Truncate image intensity
% 2. N4 bias correction: https://doi.org/10.1109/TMI.2010.2046908
% 3. SyN Registration: : https://doi.org/10.1016/j.media.2007.06.004
% 4. Warp prior to image using affine
% 5. Atropos: https://pubmed.ncbi.nlm.nih.gov/21373993/
% TODO Mention the option for single and double precision.
% TODO OASIS template citation

\subsubsection{FSL FAST}
% DOI: https://doi.org/10.1109/42.906424
% Segmentation

\subsubsection{FreeSurfer recon-all}
% Surface Reconstruction: https://doi.org/10.1006/nimg.1998.0395
% Refer to the fMRIPrep specific 3 stages workflow.

\subsubsection{FSL FLIRT}
% Registration
% FSL FLIRT: https://doi.org/10.1006/nimg.2002.1132, https://doi.org/10.1016/S1361-8415(01)00036-6, https://doi.org/10.1016/j.neuroimage.2009.06.060

\subsubsection{FSL MCFLIRT}
% Head-Motion Correction
% FSL MCFLIRT: https://doi.org/10.1006/nimg.2002.1132

\subsubsection{fMRIPrep}
\MD{TODO import fMRIPrep citation boilerplate from "./fmriprep-citations"}

@@

We used the the OpenNeuro ds004513 v1.0.2 dataset~\cite{ds004513:1.0.2}. The anatomical, functional, and diffusion data from 20 health individuals was acquired from two different cohorts. The within-subject cohort has nine participants (mean age=43~yrs, std=7~yrs; 4 females) with two session: eyes open and eyes closed. The replication cohort has eleven participants (mean age=27~yrs, std=5~yrs; 6 females) with only the eyes open session. We profiled the application using both cohort on the eyes open session, unless explicitly specified.

\subsection{Profiling}
Profiling MRI pre-processing pipelines present a set of challenges. The neuroimaging community develops pipelines using several programming languages including \textit{C}, \textit{C++}, \textit{Fortran}, \textit{Matlab}, and \textit{Python}, with several pipelines utilizing a combination of multiple languages. Pipelines are generally complex and computationally expensive.

The Intel VTune profiler solves these challenges by offering multi-language support, low performance overhead, and information of functions at runtime level. However, common profiling challenges remain. Profilers require source code to be compiled with Debug Symbols to report useful information about the different functions and modules. Moreover, different input data can substantially vary the analysis due to conditional branching in the application and convergence thresholds. We discuss our approach to these challenges in the next section.
\MD{TODO Cite effect from varying input data on numerical stability and reduced precision. Ask yohan for refereces}

First, to obtain human readable information on function and module names in VTune, we re-compiled each pipeline with debug information in a Docker image. For use on HPC systems, we created Apptainer~\cite{Kurtzer2017-bu} images using Docker2Singularity\footnote{\href{https://github.com/singularityhub/docker2singularity}{https://github.com/singularityhub/docker2singularity}} \lstlistingname~\ref{alg:vtune_example} illustrate the profiling of a pipeline by mounting the VTune binary to the Apptainer image at execution.

\begin{minipage}{\linewidth}
	\lstinputlisting[
		language={sh},
		caption={Profiling an application within a Apptainer container, using VTune profiler.},
		captionpos=b,
		label=alg:vtune_example,
		numbers=left,
		frame=single,
		basicstyle=\footnotesize,
		numbersep=5pt,
		numberstyle=\tiny\color{gray},
		rulecolor=\color{black},
		tabsize=2,
	]{vtune-example.sh}
\end{minipage}

Second, we aggregate the profiling results of each pipeline across all subjects to obtain an average performance profile. The Pareto principle is a power law which states that 80\% of the effects come from 20\% of the causes. We hypothesize that the Pareto principle applies to the performance of MRI pre-processing pipelines, which is consistent with the literature~\cite{Kukunas2015-jd}. Therefore, for fine-grained analysis, we only study the functions with the largest contribution to CPU time, up to 80\% cumulative total runtime of the application. This heuristic allows us to focus our efforts on the performance critical sections of a pipeline. We report the mean and standard deviation of the execution time for those functions.

Last, we profile the application using two threading configuration: single-threaded and 32 threads. On the one hand, the single-threaded approach simplifies the analysis and limits the potential overhead from I/O on the profiling. 
On the other hand, we use all available CPU cores from a compute node (i.e. 32) to study the multi-threading performance of the application. This is a more realistic usage for the user of the application.

@@

\subsection{Metrics Definition}
We define the makespan as the elapsed time from the start to the end of the application. The CPU time refers to the total time spent by the CPU, across all core, to execute the application. VTune provides several metrics to characterize the performance of an application. We describe the memory metrics used in this paper.

\begin{figure*}[h!]
	\noindent
	\begin{minipage}
		{0.5\textwidth}
		\begin{align*}
			C & = CPU\_CLK\_UNHALTED.THREAD                \\
			S & = CYCLE\_ACTIVITY.STALLS\_LDM\_PENDING     \\
			O & = CYCLE\_ACTIVITY.STALLS\_L1D\_PENDING     \\
			T & = CYCLE\_ACTIVITY.STALLS\_L2\_PENDING      \\
			W & = MEM\_L3\_WEIGHT                          \\
			H & = MEM\_LOAD\_UOPS\_RETIRED.LLC\_HIT        \\
			R & = MEM\_LOAD\_UOPS\_MISC\_RETIRED.LLC\_MISS \\
			M & = \frac{W \times R}{H + R}                 \\
			N & = \frac{H}{H + W \times R}                 
		\end{align*}
	\end{minipage}
	\begin{minipage}
		{0.5\textwidth}
		\begin{align}
			\%~Memory~Bound & = \frac{S}{C} \times 100          \\
			\%~L1~Bound     & = \frac{S - O}{C} \times 100      \\
			\%~L2~Bound     & = \frac{O - T}{C} \times 100      \\
			\%~L3~Bound     & = \frac{T \times N}{C} \times 100 \\
			\%~DRAM~Bound   & = \frac{T \times M}{C} \times 100 
		\end{align}
	\end{minipage}
	\caption{Intel performance monitoring events and derived memory metrics}
	\label{fig:memory-metrics}
\end{figure*}

The memory bound metrics counts the number of cycles where the CPU is starved while waiting for in-flight memory demand load. L1 bound occurs when a demand load stall some CPU cores although the data already reside in L1 cache. Similarly, L2, L3, and DRAM bound occurs when a demand load stall the CPU and there is a cache miss in the previous level of cache, but the data reside in the L2, L3, and DRAM respectively. The store bound occurs when a store operation stall the CPU. Figure~\ref{fig:memory-metrics} shows the monitoring events and equation to derive the metrics defining the memory boudnd characterization~\cite{Intel2006-lc}. We use the convention from~\cite{Kukunas2015-jd} for the equations.

The difference between the memory bound value and the sum of the other metrics represent the percentage of CPU starvation while waiting from data absent from all cache level; i.e. the data is fetched from disk.

\subsection{Infrastructure}
For our experiments, we used the \textit{Slashbin cluster} at Concordia University. The compute nodes are configured with two 16-core Intel(R) Xeon(R) Gold 6130 CPU @ \SI{2.10}{\giga\hertz}, \SI{250}{\gibi\byte} of RAM, \SI{126}{\gibi\byte} of tmpfs, six \SI{447}{\gibi\byte} SSDs with the XFS file system (no RAID enabled), Rocky~Linux~8.9 and Linux kernel \textit{4.18.0-477.10.1.el8\_lustre.x86\_64}.

\section{Results}
In this section we present the aggregated profiling data obtained with VTune. First, we present a performance overview of the pipelines. We then dive into more detailed analysis observed from the hotspots of the pipelines. Unless specified explicitly, we only present the profile data from using 1-thread since the results from 32-threads are similar.

\subsection{Long-tail Distribution}
Figure~\ref{fig:long-tail-distribution} shows that 80\% of the total CPU time is spent in less than 1.5\% of the function across all pipelines. While consistent with the Pareto principle, the distribution of the functions' CPU time does not follow a pareto distribution. We plotted the distribution of the functions' CPU time on a log-log scale and found it to be closer to a Zipf's law distribution with exponent $\alpha=1.8$.

\begin{figure}[h]
	\centering
	\includegraphics[width=\linewidth]{figures/global-longtail.png}
	\caption{Distribution of the functions' CPU time. The left y-axis shows the average total CPU Time spent in a function, while the right y-axis show the cumulative CPU time percentage. The x-axis percentage of functions ordered by decreasing CPU time. The data includes all functions from all pipelines.}
	\label{fig:long-tail-distribution}
\end{figure}

\subsection{Makespan: A performance overview}
Figure~\ref{fig:makespan-1thread} depicts the total time elapsed from the start to the end of each pipeline; i.e. makspan. There is a significant difference between the makespan of different applications. This is expected as the pipelines perform different tasks. The standard deviation is nearly null for all pipelines, except for FreeSurfer recon-all which as a small variation. Overall, this shows that the input data does not substantially affect the makespan of those pipelines.

\begin{figure}[t]
	\centering
	\includegraphics[width=\linewidth]{figures/makespan-1thread.png}
	\caption{Average makespan of the pipeline while using 1-thread. The error bars show the standard deviation. \TG{Font size in the figure is too small, it should be similar to the caption font size}}
	\label{fig:makespan-1thread}
\end{figure}

\subsection{Impact of Memory Bounded Functions}
Table~\ref{extab:memory-bound} shows that fetching data from disk is the main source of cycles where the CPU is starved. Following, The L1 bound has the highest cache-level contribution to the memory bound of the pipelines. ANTS pipelines have the highest L1 bound, followed by FreeSurfer recon-all and FSL MCFLIRT. L2 and L3 bound is low for all pipeline, although highest for FreeSurfer recon-all. The DRAM bound is high for both FSL FLIRT and FreeSurfer recon-all, followed by ANTS brainExtraction in single and double precision. Store bound is low for all pipelines. We note an interesting relationship between the low L1 bound and the high DRAM bound of FreeSurfer recon-all and FSL FLIRT, compared to the other pipelines.

\csvnames{csvcol}{1=\pipeline, 2=\mem, 3=\la, 4=\lb, 5=\lc, 6=\dram, 7=\store}
\csvstyle{memory_bound}{
	tabular = |r|c|c|c|c|c|c|,
	table head = \hline Pipeline &  \% Memory Bound &  \% L1 Bound &  \% L2 Bound &  \% L3 Bound & \% DRAM Bound & \% Store Bound\\\hline\hline,
	late after line = \\\hline,
	respect all,
	csvcol
}
\begin{table*}[ht]
	\centering
	\csvreader[memory_bound]{tables/memory_bound.csv}{}
	{\pipeline & \tablenum[round-precision=2]{\mem} & \tablenum[round-precision=2]{\la} & \tablenum[round-precision=2]{\lb} & \tablenum[round-precision=2]{\lc} & \tablenum[round-precision=2]{\dram} & \tablenum[round-precision=2]{\store}}
	\caption{Impact of data load on stalled CPU cycles. The values are the summation of the metric weighted by each function CPU time. This represent the percentage of the total CPU time stalled by each metrics. We show the average value across all subjects execution with one thread.}
	\label{extab:memory-bound}
\end{table*}

\subsection{Main Bottleneck: Linear Interpolation}
Table~\ref{extab:interpolation} shows that interpolation (or convolution) is a critical part for all pipelines, except FreeSurfer recon-all. For those pipelines, interpolation contributes between 32.20\% to 62.70\% of the total CPU time. Overall, the number of functions using interpolation is low with less than 20 in each applications. In other words, few interpolation functions are contribute to a large portion of the CPU time.
% There seems to be few  the stages from recon-all: https://surfer.nmr.mgh.harvard.edu/fswiki/recon-all (see section "Manual-Intervention Workflow Directives": stages 1-31)

\csvnames{csvcol}{1=\pipeline, 2=\nfunc, 3=\cputime}
\csvstyle{interpolation}{
	tabular = |r|c|c|,
	table head = \hline Pipeline & \# of functions & \% CPU Time \\
	& with interpolation & from interpolation\\\hline\hline,
	late after line = \\\hline,
	respect all,
	csvcol
}
\begin{table*}[h]
	\centering
	\csvreader[interpolation]{tables/interpolation.csv}{}
	{\pipeline & \nfunc & \tablenum[round-precision=2]{\cputime}}
	\caption{Contribution of interpolation to the applications' total CPU time. The percentage is the average sum of CPU time of functions using interpolation. The data includes all functions; not only the top 80\% of the CPU time.}
	\label{extab:interpolation}
\end{table*}
